---
title: "Week 14 Group Activity Handout"
format:
  pdf:
    documentclass: article
    geometry: margin=0.8in
    toc: false
    number-sections: false
---

```{r}
#| include: false

library(ggplot2)

accent <- "#C8C372"

handout_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank(),
    legend.title = element_blank(),
    legend.background = element_rect(fill = "white", color = NA),
    plot.title.position = "plot",
    axis.title.x = element_text(margin = margin(t = 8)),
    axis.title.y = element_text(margin = margin(r = 8))
  )
```

# Visual Redesign Activity

Work in pairs or groups of three. For each visual:

1. Identify the audience.
2. Identify the decision.
3. Identify the goal of the visual.
4. Decide what to remove, mute, or highlight.
5. Rewrite the title so the takeaway is obvious.

:::{.callout-note}
Prompt template

Use this sentence frame before you redesign any chart:

The audience is `______`. They need to decide `______`. The goal of the visual is to `______`.
:::

\newpage

## Activity Visual A

Scenario:

- Audience: warehouse operations leadership
- Decision: which warehouse to prioritize for training and process improvement
- Goal: show which location is underperforming based on on-time delivery

What would you remove, mute, highlight, and rewrite?

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6.8
#| fig-height: 4.4

activity_a <- data.frame(
  warehouse = rep(c("Fort Collins", "Denver Metro", "Pueblo", "Grand Junction"), each = 4),
  quarter = rep(c("Q1", "Q2", "Q3", "Q4"), times = 4),
  rate = c(96, 95, 94, 95, 93, 91, 92, 90, 88, 94, 93, 92, 91, 90, 89, 88)
)

ggplot(activity_a, aes(x = warehouse, y = rate, fill = quarter)) +
  geom_col(position = position_dodge(width = 0.82), width = 0.7, color = "gray35") +
  geom_text(
    aes(label = paste0(rate, "%")),
    position = position_dodge(width = 0.82),
    vjust = -0.35,
    size = 3.2
  ) +
  scale_fill_manual(values = c("Q1" = "#4F81BD", "Q2" = "#9BBB59", "Q3" = "#C0504D", "Q4" = "#8064A2")) +
  scale_y_continuous(
    breaks = seq(80, 100, 2),
    expand = expansion(mult = c(0, 0.06))
  ) +
  coord_cartesian(ylim = c(80, 100)) +
  labs(
    title = "On-Time Delivery Rate by Warehouse and Quarter",
    x = "Warehouse",
    y = "On-time delivery (%)"
  ) +
  handout_theme +
  theme(
    panel.grid.major = element_line(color = "gray65", linewidth = 0.7),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    panel.border = element_rect(color = "gray35", fill = NA, linewidth = 0.8)
  )
```

:::{.callout-tip}
Redesign checklist

- What should the audience notice first?
- Which details can be removed without losing meaning?
- Which parts should be pushed into the background?
- How would you rewrite the title so it states the takeaway?
:::

\newpage

## Activity Visual B

Scenario:

- Audience: sales operations leadership
- Decision: which customer acquisition channels to invest in
- Goal: compare the cost-effectiveness of different customer acquisition channels

What would you remove, mute, highlight, and rewrite?

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6.8
#| fig-height: 4.4

activity_b <- data.frame(
  year = rep(2019:2024, 4),
  channel = rep(c("Search", "Email", "Social", "Referral"), each = 6),
  cost = c(42, 44, 45, 46, 47, 49,
           38, 37, 36, 35, 35, 34,
           31, 34, 38, 43, 48, 52,
           45, 44, 43, 43, 42, 42)
)

ggplot(activity_b, aes(x = year, y = cost, color = channel, group = channel)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 2.8) +
  scale_color_manual(
    values = c(
      "Search" = "#4F81BD",
      "Email" = "#9BBB59",
      "Social" = "#C0504D",
      "Referral" = "#8064A2"
    )
  ) +
  scale_x_continuous(breaks = 2019:2024) +
  scale_y_continuous(
    limits = c(28, 55),
    breaks = seq(30, 55, 5),
    expand = expansion(mult = c(0, 0.03))
  ) +
  labs(
    title = "Customer Acquisition Cost by Channel",
    x = "Year",
    y = "Cost per acquired customer ($)"
  ) +
  handout_theme +
  theme(
    panel.grid.major = element_line(color = "gray65", linewidth = 0.7),
    legend.position = "bottom",
    panel.border = element_rect(color = "gray35", fill = NA, linewidth = 0.8)
  )
```

:::{.callout-tip}
Redesign checklist

- What should a decision-maker notice first?
- Would you keep one chart or split the story?
- Which line should be emphasized?
- How would you rewrite the title to make the message immediate?
:::

## Debrief

Be ready to give a 30-second redesign pitch.

- Who is the audience?
- What decision depends on the visual?
- What is the visual goal?
- What did you remove, mute, or highlight?
- What takeaway should the audience see before anything else?
