---
title: "Week 7: Regression as a Tool for Answering Questions"
format:
  revealjs:
    slide-number: true
    theme: [night, ../custom.scss]
    touch: true
---

## 

<!-- clicker questions -->

## Motivating Questions
:::{.incremental}
- Do fuel stations closer to highways charge higher prices?
- Does advertising increase sales?
- Do higher wages lead to higher productivity?
:::


## Running example: <br>Corn Yield and Drought

What is the effect of drought on corn yields?

<br>

```{mermaid}
%%{init: {"themeVariables": {"fontSize": "24px"}}}%%
flowchart LR
  G[Goals] ==> P[Problem]
  P ==> Q[Question]
  Q ==> Da[Data]
  Da ==> M[Model]
  M ==> R[Result]
  R ==> D[Decision]

  style M fill:#FFD966,stroke:#333,stroke-width:3px,color:#000
```

<br>

We seek a method that tells us how much corn yields change when drought conditions change.

## 
:::{.main-point}
Regression is a method for understanding relationships between quantitative variables
:::

## Example: Gas prices and highway proximity

- Do fuel stations closer to highways charge higher prices?
- Can you tell by looking at a map of gas stations and highways?
- <https://www.gasbuddy.com/gasprices/wyoming/cheyenne>

## How does regression work?

- Regression estimates the relationship between a dependent variable and one or more independent variables.
- It quantifies how much the dependent variable changes when the independent variable(s) change.
- It does this by "fitting a line" through the data points that best captures the relationship.
- What do we mean by "best" fit?

## Regression demo

- What is the equation of a line?
- The data give us our x and y values. Coordinates.

Go to R

:::{.notes}
- motivate the generation of the data
- experiment with the class moving the line
- show how the line changes with different slopes and intercepts (parameters)
- the solution is the line that minimizes the sum of squared errors. why squared errors? why not absolute errors?

:::

## Regression math{.scrollable}

- The regression line is defined by the equation: 
$$
y = \alpha + \beta x + \epsilon
$$

    - $\alpha$ is the intercept (the value of $y$ when $x=0$)
    - $\beta$ is the slope (the change in $y$ for a one-unit change
    in $x$)
    - $\epsilon$ is the error term (the difference between the observed and predicted values)

Rearranging gives us:
$$
y - \alpha - \beta x = \epsilon
$$

The regression line is the one that minimizes the sum of squared errors:
$$
\min_{\alpha, \beta} \sum_{i=1}^n (y_i - \alpha - \beta x_i)^2
$$

Differentiating with respect to $\alpha$ and $\beta$ and setting the derivatives to zero gives us the normal equations, which we can solve to find the estimates of $\alpha$ and $\beta$.

$$
\begin{cases}
\sum_{i=1}^n (y_i - \alpha - \beta x_i) = 0 \\
\sum_{i=1}^n x_i (y_i - \alpha - \beta x_i) = 0
\end{cases}
$$

We have 2 equations with 2 unknowns. Solving these equations yields the least squares estimates:
$$
\hat{\beta} = \frac{\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^n (x_i - \bar{x})^2}
$$
$$
\hat{\alpha} = \bar{y} - \hat{\beta} \bar{x}
$$

$\beta$ is also the covariance of $x$ and $y$ divided by the variance of $x$:
$$
\hat{\beta} = \frac{\text{Cov}(x, y)}{\text{Var}(x)}
$$  

# Interpretting Regression Output

##
- The regression output provides estimates of the coefficients (intercept and slope) along with their standard errors, t-values, and p-values.

## Example regression output:

<br>

| Variable | Estimate | Std. Error | t-stat | p-value |
|----------|:--------:|:----------:|:------:|:-------:|
| x        | -4.23    | 0.23       | -18.39 | <0.001  |
| Intercept | 4.48     | 0.27       | 16.59  | <0.001  |


## Interpreting Coefficient Estimates

- The coefficient estimates tell us the **expected change** in the dependent variable for a one-unit change in the independent variable
- In our example, the slope of -4.23 means that for every one-unit increase in x, y is expected to decrease by 4.23 units.
- If x was miles from the highway, this would mean that for every additional mile from the highway, gas prices are expected to decrease by 4.23 cents.

## Estimating Uncertainty: Standard Errors

- The estimates are just that: estimates. They are subject to sampling variability.
- What if we repeated the study with a different sample of gas stations? or a different period in time? Would we get the same estimates?
- The standard error gives us a measure of how much the estimate would vary across different samples.

## Hypothesis Testing{.scrollable}

- We make assumptions about the error term $\epsilon$ (e.g., it is normally distributed with mean 0 and constant variance) that allow us to perform hypothesis tests on the coefficients.
- The null hypothesis is typically that the coefficient is equal to zero (no effect).
- The t-statistic is the estimate divided by its standard error, and the p-value tells us the probability of observing such an extreme t-statistic if the null hypothesis were true.
- In our example, the t-statistic of -18.39 for the slope indicates that the coefficient is highly statistically significant (p < 0.001), suggesting a strong relationship between distance from the highway and gas prices.

:::{.notes}
Imagine testing whether a coin is fair.
If you flip 10 times and get 6 heads:
Not that surprising.
If you flip 100 times and get 95 heads:
Very surprising if coin is fair.
The p-value quantifies “how surprising.”
:::


## 
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4.2


df <- 52
t_obs <- -18.39
alpha <- 0.05
t_crit <- qt(1 - alpha/2, df)

# Plot range chosen for teaching: shows critical values clearly
x <- seq(-5, 5, length.out = 2000)
y <- dt(x, df)

plot(x, y, type = "l", xlab = "t", ylab = "Density",
     main = sprintf("t distribution (df = %d): two-sided test at α = %.2f", df, alpha))

# Shade rejection regions (tails beyond ±t_crit)
x_left  <- x[x <= -t_crit]
y_left  <- dt(x_left, df)
polygon(c(x_left, rev(x_left)), c(y_left, rep(0, length(y_left))),
        col = "gray80", border = NA)

x_right <- x[x >=  t_crit]
y_right <- dt(x_right, df)
polygon(c(x_right, rev(x_right)), c(y_right, rep(0, length(y_right))),
        col = "gray80", border = NA)

# Draw critical value lines
abline(v = c(-t_crit, t_crit), lty = 2)
text(-t_crit, max(y)*0.95, sprintf("-t* = %.2f", t_crit), pos = 2)
text( t_crit, max(y)*0.95, sprintf("t* = %.2f", t_crit), pos = 4)



```

$t=(\beta - 0)/\sigma_\beta$

## 
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4.2

df <- 52
t_obs <- -18.39
alpha <- 0.05
t_crit <- qt(1 - alpha/2, df)

# Plot range chosen for teaching: shows critical values clearly
x <- seq(-5, 5, length.out = 2000)
y <- dt(x, df)

plot(x, y, type = "l", xlab = "t", ylab = "Density",
     main = sprintf("t distribution (df = %d): two-sided test at α = %.2f", df, alpha))

# Shade rejection regions (tails beyond ±t_crit)
x_left  <- x[x <= -t_crit]
y_left  <- dt(x_left, df)
polygon(c(x_left, rev(x_left)), c(y_left, rep(0, length(y_left))),
        col = "gray80", border = NA)

x_right <- x[x >=  t_crit]
y_right <- dt(x_right, df)
polygon(c(x_right, rev(x_right)), c(y_right, rep(0, length(y_right))),
        col = "gray80", border = NA)

# Draw critical value lines
abline(v = c(-t_crit, t_crit), lty = 2)
text(-t_crit, max(y)*0.95, sprintf("-t* = %.2f", t_crit), pos = 2)
text( t_crit, max(y)*0.95, sprintf("t* = %.2f", t_crit), pos = 4)

# Mark observed t.
# It's off the plotted range; show an arrow indicating it's far beyond the left tail.
if (t_obs < min(x)) {
  arrows(x0 = -4.2, y0 = dt(-4.2, df)+.03, x1 = -4.95, y1 = dt(-4.95, df) +.03, length = 0.1)
  text(-4.2, dt(-4.2, df)+.03, sprintf("observed t = %.2f (far left)", t_obs), pos = 4)
} else {
  abline(v = t_obs, lwd = 2)
  text(t_obs, max(y)*0.8, sprintf("observed t = %.2f", t_obs), pos = 4)
}

# Optional: label acceptance region
#text(0, max(y)*0.35, "Fail to reject H₀ region", cex = 0.9)
#text(-3.2, dt(-3.2, df)*1.2, "Reject H₀", cex = 0.9)
#text( 3.2, dt( 3.2, df)*1.2, "Reject H₀", cex = 0.9)
```

## {.iclicker} 
<div style="font-size: 0.75em;">
Dependent variable: Gas Price ($/gallon)

| Variable                    | Estimate | Std. Error | t-stat | p-value |
| --------------------- | -------- | ---------- | ------ | --------- |
| Distance to Highway (miles) | -0.04    | 0.02       | -2.00  | 0.048   |
| Constant                    | 3.50     | 0.10       | 35.00  | <0.001  |


:::{.question}
What does the coefficient on distance to highway mean?
:::

For every additional mile from the highway...

:::{.choices}
A. gas prices are expected to decrease by 4 cents.

B. gas prices are expected to decrease by 0.02 cents.

C. gas prices are expected to increase by 3.50 dollars.

D. gas prices are expected to decrease by 4 dollars.
:::
</div>

## {.iclicker} 
<div style="font-size: 0.65em;">
Dependent variable: Daily Mountain Bike Trail Visits

| Variable          | Estimate | Std. Error | t-stat | p-value |
| ----------------- | -------- | ---------- | ------ | ------- |
| Temperature (°F)  | 12.5     | 8.3        | 1.51   | 0.130  |
| Constant          | -300.0   | 150.0      | -2.00  | 0.048   |

:::{.question}
Which of the following is true about the relationship between temperature and daily mountain bike trail visits?
:::

:::{.choices}
A. Each 1°F increase in temperature causes 12.5 more trail visits per day.

B. A 1°F increase in temperature is associated with 12.5 more visits per day, and this relationship is statistically significant at the 5% level.

C. A 1°F increase in temperature is associated with 12.5 more visits per day, but the estimate is not statistically significant at the 5% level.

D. Temperature has no relationship with trail visits because the p-value is greater than 0.05.
:::
</div>


## Summary

- Regression is a powerful tool for quantifying relationships between variables.
- Coefficient estimates tell us the expected change in the dependent variable for a one-unit change in the independent variable.
- Standard errors and hypothesis tests help us understand the uncertainty around our estimates.
- Interpretation of regression results requires careful consideration of the context and the assumptions underlying the model.

## Lab Preview

- We will use regression to analyze the relationship between drought conditions and corn yields.
- We will learn how to fit a regression model, interpret the output, and evaluate the model's assumptions and performance.